---
title: "Activity 1.1 - vector basics"
format: pdf
editor_options: 
  chunk_output_type: console
---

***SUBMISSION INSTRUCTIONS***: 

1. Complete this activity, rendering periodically to html to view your output.  
2. When complete, render to pdf:
  a. Click on the **Terminal** window (next to **Console**) and type `quarto install tinytex` then hit *Enter*
  b. In the `yaml` at the top of this document (the settings inside the `---` at the top), change the format to `format: pdf`
  c. Render the document, make sure a pdf is created
3. Submit **both** the .pdf and the .qmd of this file to D2L.

# Question 1

Consider the 3-dimensional vector $w = c(-5, 2, -3)$.

## A)

Find $||w||_1$, the L1 (aka taxicab or Manhattan) norm of this vector. Show your work!

## B)

Find $||w||_2$, the L2 (aka Euclidean) norm of this vector. Show your work!

## C)

Create the vector in `R`. Use `R` to verify your computations above.

# Question 2

Consider the `iris` data set, which comes packaged with a standard `R` installation:

```{r}
data(iris)
head(iris)
```

## A)

Select only the sepal and petal variables. How many vectors are there, and what is the dimension of each vector?

## B)

Find the mean and standard deviation vectors.

## C)

Mean-center and scale the data set. Verify that the mean and standard deviation vectors of the scaled data set are $\vec 0$ and $\vec 1$, respectively.

## D)

Find the centered/scaled vectors for the first two irises. Find the L1 and L2 distances between these two irises "from scratch," then verify your answer using `dist`.

# Question 3

Reconsider the `USairpollution` data:

```{r}
library(HSAUR2)
data("USairpollution")
```

## A)

GOAL: Find the cities that are most and least similar with respect to their pollution. Use `dist` to find L2 distances between the cities. Then use wrangling approaches to find the two cities that are most similar, and two cities that are most dissimilar.

## B)

Assess how scaling the data before computing distance impacts your answer to the previous question.

## C)

Which city is Minneapolis most similar to? Most dissimilar to?
